---
title: "Week11_12_Part1_Introduction to Machine Learning KNN"
author: "Janine Par"
date: '2022-05-31'
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(plyr)
library(ggplot2)
library (lm.beta)
library (car)
library(dplyr)
library(caret)
library(class)
library(caTools)



options(scipen=5)

theme_set(theme_minimal())
```

```{r Load files and Normalize }

#Fit a Logistic Regression Model

## Set the working directory to the root of your DSC 520 directory
setwd("C:/Users/janin/OneDrive/Documents/R_repo/dsc520/")

## Load the `data/binaryclassifierdata`
binaryclass_df <- read.csv("data/binary-classifier-data.csv", header=TRUE, comment.char = "@")

## Load the `data/Trinary Classifier Data`
trinaryclass_df <- read.csv("data/trinary-classifier-data.csv", header=TRUE, comment.char = "@")

str(binaryclass_df)

str(trinaryclass_df)

table(binaryclass_df$label)

table(trinaryclass_df$label)

#Plot the data from each dataset using a scatter plot.

ggplot(binaryclass_df, aes(x=x, y=y, color=label)) + geom_point() + geom_smooth(method="lm")

ggplot(trinaryclass_df, aes(x=x, y=y, color=label)) + geom_point() + geom_smooth(method="lm")

normalize <- function(x)  {(x -min(x))/(max(x)-min(x))}

binaryclass_df_n <- as.data.frame(lapply(binaryclass_df[2:3], normalize))  #Normalize X and Y 

trinaryclass_df_n <- as.data.frame(lapply(trinaryclass_df[2:3], normalize))  #Normalize X and Y 

```


#Work on Binary Class datasets 


```{r Binary Dataset}
#Split data

set.seed(1234)

#Get random numbers for Training Data
size <- floor(0.7*nrow(binaryclass_df))

train_ind <- sample(seq_len(nrow(binaryclass_df)), size = size)

bcsplit_train.label <- binaryclass_df[train_ind,1 ]

bcsplit_test.label <- binaryclass_df[-train_ind,1 ]

bcsplit_train <- binaryclass_df_n[train_ind,]

bcsplit_test <- binaryclass_df_n[-train_ind,]

bcsplit_predict <- knn(train = bcsplit_train,
                       test=bcsplit_test,
                       cl=bcsplit_train.label,
                       k=round(sqrt(nrow(bcsplit_train)))) # K-32 initial run 

acc_bcsplit_32 <- 100 * sum(bcsplit_test.label == bcsplit_predict)/NROW(bcsplit_test.label)

#Confusion Matrix
confmatrix <- table(Actual_value=bcsplit_test.label, Predicted_Value= bcsplit_predict)

# Accuracy 
(confmatrix [[1,1]] + confmatrix [[2,2]])/sum(confmatrix) * 100

knn_val <- c(3,5,10,15,20,25)
 
bcs_knn_model <- data.frame()

i=1

for (kv in knn_val)
{
  bcsplit_predict<- knn(train = bcsplit_train,
                         test=bcsplit_test,
                         cl=bcsplit_train.label,
                         k=kv)

  confmatrix <- table(Actual_value=bcsplit_test.label, Predicted_Value= bcsplit_predict)
  kvaccuracy <-   100 * sum(bcsplit_test.label == bcsplit_predict)/NROW(bcsplit_test.label)  #((confmatrix [[1,1]] + confmatrix [[2,2]])/sum(confmatrix))*100
  kvalue <- kv
  bcs_knn_model <- rbind(bcs_knn_model, c(kvalue, kvaccuracy))
  names(bcs_knn_model) <- c("Kvalue", "Accuracy")
}

#ggplot(data=bcs_knn_model, aes(x=kvalue, y=accuracy)) + geom_point() 

plot(bcs_knn_model)

# K value and Accuracy
bcs_knn_model

#Binary Class Linear Classifier from last week (week 10)
  
binaryclass.model <- glm(label~x+y, data=binaryclass_df, family=binomial())

summary (binaryclass.model)

bcsplit <- sample.split(binaryclass_df, SplitRatio = 0.8)

bcsplit_train <- subset(binaryclass_df,tssplit='True')

bcsplit_test <- subset(binaryclass_df,tssplit='False')

#Predict 
res.train <- predict(binaryclass.model,bcsplit_train,type ="response")

res.test <- predict(binaryclass.model,bcsplit_test,type ="response")

confmatrix <- table(Actual_value=bcsplit_train$label, Predicted_Value= res.train > 0.5)

(confmatrix [[1,1]] + confmatrix [[2,2]])/sum(confmatrix)*100

```


#Work on Trinary Class datasets 


```{r Trinary Dataset}
#Split data

set.seed(1234)

#Get random numbers for Training Data
size <- floor(0.7*nrow(trinaryclass_df))

train_ind <- sample(seq_len(nrow(trinaryclass_df)), size = size)

trsplit_train.label <- trinaryclass_df[train_ind,1 ]

trsplit_test.label <- trinaryclass_df[-train_ind,1 ]

trsplit_train <- trinaryclass_df_n[train_ind, ]

trsplit_test <- trinaryclass_df_n[-train_ind,]

trsplit_predict <- knn(train = trsplit_train,
                       test=trsplit_test,
                       cl=trsplit_train.label,
                       k=round(sqrt(nrow(trsplit_train)))) # K-32 initial run 

summary(trsplit_predict)


#COnfusion Matrix
confmatrix <- table(Actual_value=trsplit_test.label, Predicted_Value= trsplit_predict)

confmatrix

# Accuracy 
(confmatrix [[1,1]] + confmatrix [[2,2]])/sum(confmatrix) * 100


#Fit a k nearest neighbors’ model for each dataset for k=3, k=5, k=10, k=15, k=20, and k=25. Compute the accuracy of the resulting models for each value of k. Plot the results in a graph where the x-axis is the different values of k and the y-axis is the accuracy of the model.

knn_val <- c(3,5,10,15,20,25)

tr_knn_model <- data.frame()

i=1

for (kv in knn_val)
{
  trsplit_predict<- knn(train = trsplit_train,
                        test=trsplit_test,
                        cl=trsplit_train.label,
                        k=kv)
  
  confmatrix <- table(Actual_value=trsplit_test.label, Predicted_Value= trsplit_predict)
  kvaccuracy <-   100 * sum(trsplit_test.label == trsplit_predict)/NROW(trsplit_test.label)  
  kvalue <- kv
  tr_knn_model <- rbind(tr_knn_model, c(kvalue, kvaccuracy))
  names(tr_knn_model) <- c("Kvalue", "Accuracy")
}

#ggplot(data=bcs_knn_model, aes(x=kvalue, y=accuracy)) + geom_point() 

plot(tr_knn_model)

tr_knn_model

```


#Looking back at the plots of the data, do you think a linear classifier would work well on these datasets?

**I do not think that the linear classifier will work well with the Binary and Trinary dataset because the plot shows non-linearity of the data.** 

#How does the accuracy of your logistic regression classifier from last week compare?  Why is the accuracy different between these two methods?

**Last week, Binary dataset has lower accuracy of 58.3% compare to KNN model which resulted to higher accuracy and this is because KNN model fits the data more because of it's Non-Linearity.** 
  